### 2016-01-11 papers/CAFe/HIPS-paper/coarray-extensions.tex
Reduced font size of verbatim text.

A programmer requests a subimage by executing the new CAFe function,
\begin{verbatim}
   device = GET_SUBIMAGE(device_id)
\end{verbatim}
where the integer argument represents an attached hardware device (or perhaps a separate
process or a team of threads; it is the compilers responsibility to determine precisely
what a subimage means in terms of the software and hardware environment under its
control).  If the function fails (e.g., the requested device is unavailable) it returns
the local image number of the process that is executing the current program, obtained by a
call to the Fortran 2008 function \texttt{this\_image()}.  Returning the current image if
the \texttt{GET\_SUBIMAGE} fails, allows program execution to proceed correctly even if
there are no attached devices.

@@

In Fortran, arrays that are visible to other program images must be declared with the
codimension attribute, for example,
\begin{verbatim}
 real, allocatable :: U(:,:)[:], V(:,:)[:]
\end{verbatim}
declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
corank of 1.  In Fortran, square brackets \texttt{[ ]} are used to indicate operations
(possibly expensive) on distributed memory where parentheses \texttt{( )} are used to
select specific array elements; square brackets, on the other hand, denote the
\emph{location} of the array or array elements selected.

@@

As discussed, CAFe allows coarray memory to be conditionally allocated on a subimage.
It should be allocated conditionally because the requested subimage in a \texttt{GET\_SUBIMAGE}
call may not be available.  In addition, not all coarrays need be allocated on each
subimage.  For example, the following code segment allocates memory for coarrays
\texttt{U} and \texttt{V} on all images, but only coarray \texttt{U} is allocated on
the subimage specified by \texttt{device}:
\begin{verbatim}
   allocate(U(M,N)[*], V(M,N)[*])
   if (device \= this_image()) then
     allocate(U(M,N)[*])  [[device]]
   end if
\end{verbatim}
where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement
allocates coarrays \texttt{U} and \texttt{V} on all program images.  The \texttt{*} symbol
in the allocation is required for the last corank dimension in CAF (in this case there is
only one) to allow for the number of program images to be a runtime parameter, which can
be ascertained with a \texttt{num\_images()} function call.  If \texttt{device} is
available, its numeric value will be different from \texttt{this\_image()} and thus
coarray \texttt{U} will also be allocated on the device.  Note that placement of the
memory is specified by the use of double square bracket notation \texttt{[[ ]]}.  Like
single square brackets, double square brackets indicate something special (and possibly
expensive) is to occur, in this case possibly remote memory allocation.

Deallocation of subimage memory is similar to allocation,
\begin{verbatim}
   if (device \= this_image()) then
     deallocate(U)  [[device]]
   end if
\end{verbatim}

@@

Once memory is allocated on the device, it can be initialized by copying memory from the
hosting image to the device.  This is explicitly done with normal CAF syntax.  For example,
\begin{verbatim}
  U[device] = 3.14
\end{verbatim}
assigns 3.14 to all of the array elements of \texttt{U} located on \texttt{device}.
Specific array elements can also be transferred, 
\begin{verbatim}
  U(1,:) = U(1,:)[device]
  U(M,:) = U(M,:)[device]
\end{verbatim}
where the first and last columns of \texttt{U} are copied from the device to the hosting image.

@@

However, CAFe allows images to execute tasks on hosted subimages using standard procedure
calls.  For example,
\begin{verbatim}
  call relax(U[device])  [[device]]
\end{verbatim}
executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
whereby the double square bracket notation indicates \emph{where} the function will execute.
Funtions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the
\texttt{value} attribute).

@@

In addition, CAFe subimage tasks may also be defined by the body of a Fortran do
concurrent construct and executed on a subimage.  For example, the code segment,
\begin{verbatim}
do concurrent(i=1:N-1)  [[device]]
 begin block
   real :: w = 0.67
   S(i) = (1-w)*T(i) + w*(T(i-1)+T(i+1)/2)
 end block
end do
\end{verbatim}
will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.
Execution of a \texttt{do concurrent} loop on a subimage acts \emph{as if} the loop were
extracted as a outlined function and the subsequent task executed on a subimage.  Thus
there is an implicit barrier at the end of the loop that blocks the hosting image
from continuing execution until \emph{all} of the iterates of the loop have finished.

@@

Thus we introduce another CAFe function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
variable representing a fully-formed (established) team consisting of only one team
member, the current subimage.  If this function is called outside the context of a task
executing on a subimage it will return only a team consisting of the current image,
\texttt{this\_image()}.  Synchronization across possible subimage threads is then possible
using the statement
\begin{verbatim}
  sync team (GET_SUBIMAGE_TEAM())
\end{verbatim}
This allows implementation of the technique of double buffering within a subimage task
so that all of the \emph{potential} subimage threads will have completed operating on the
temporary buffer before it is copied back to the primary array variable.

Fortran 2015 also introduces the concept of events.  An event is a Fortran type that
can be posted to by one image and waited for by another image.  CAFe relaxes the
constraint that events must be coarrays so that events can be used locally within
a single image to control concurrent execution of multiple subimage tasks.  For example,
the following code segment shows how two tasks can be run concurrently on two separate
subimages:
\begin{verbatim}
 type(event_type) :: evt
 call update_ocn [[dev1, WITH_EVENT=evt]]
 call update_atm [[dev2, WITH_EVENT=evt]]
 event wait (evt, until_count=2)
\end{verbatim}

>>>>

A programmer requests a subimage by executing the new CAFe function,
\small
\begin{verbatim}
   device = GET_SUBIMAGE(device_id)
\end{verbatim}
\normalsize
where the integer argument represents an attached hardware device (or perhaps a separate
process or a team of threads; it is the compilers responsibility to determine precisely
what a subimage means in terms of the software and hardware environment under its
control).  If the function fails (e.g., the requested device is unavailable) it returns
the local image number of the process that is executing the current program, obtained by a
call to the Fortran 2008 function \texttt{this\_image()}.  Returning the current image if
the \texttt{GET\_SUBIMAGE} fails, allows program execution to proceed correctly even if
there are no attached devices.

@@

In Fortran, arrays that are visible to other program images must be declared with the
codimension attribute, for example,
\small
\begin{verbatim}
 real, allocatable :: U(:,:)[:], V(:,:)[:]
\end{verbatim}
\normalsize
declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
corank of 1.  In Fortran, square brackets \texttt{[ ]} are used to indicate operations
(possibly expensive) on distributed memory where parentheses \texttt{( )} are used to
select specific array elements; square brackets, on the other hand, denote the
\emph{location} of the array or array elements selected.

@@

As discussed, CAFe allows coarray memory to be conditionally allocated on a subimage.
It should be allocated conditionally because the requested subimage in a \texttt{GET\_SUBIMAGE}
call may not be available.  In addition, not all coarrays need be allocated on each
subimage.  For example, the following code segment allocates memory for coarrays
\texttt{U} and \texttt{V} on all images, but only coarray \texttt{U} is allocated on
the subimage specified by \texttt{device}:
\small
\begin{verbatim}
   allocate(U(M,N)[*], V(M,N)[*])
   if (device \= this_image()) then
     allocate(U(M,N)[*])  [[device]]
   end if
\end{verbatim}
\normalsize
where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement
allocates coarrays \texttt{U} and \texttt{V} on all program images.  The \texttt{*} symbol
in the allocation is required for the last corank dimension in CAF (in this case there is
only one) to allow for the number of program images to be a runtime parameter, which can
be ascertained with a \texttt{num\_images()} function call.  If \texttt{device} is
available, its numeric value will be different from \texttt{this\_image()} and thus
coarray \texttt{U} will also be allocated on the device.  Note that placement of the
memory is specified by the use of double square bracket notation \texttt{[[ ]]}.  Like
single square brackets, double square brackets indicate something special (and possibly
expensive) is to occur, in this case possibly remote memory allocation.

Deallocation of subimage memory is similar to allocation,
\small
\begin{verbatim}
   if (device \= this_image()) then
     deallocate(U)  [[device]]
   end if
\end{verbatim}
\normalsize

@@

Once memory is allocated on the device, it can be initialized by copying memory from the
hosting image to the device.  This is explicitly done with normal CAF syntax.  For example,
\small
\begin{verbatim}
  U[device] = 3.14
\end{verbatim}
\normalsize
assigns 3.14 to all of the array elements of \texttt{U} located on \texttt{device}.
Specific array elements can also be transferred, 
\small
\begin{verbatim}
  U(1,:) = U(1,:)[device]
  U(M,:) = U(M,:)[device]
\end{verbatim}
\normalsize
where the first and last columns of \texttt{U} are copied from the device to the hosting image.

@@

However, CAFe allows images to execute tasks on hosted subimages using standard procedure
calls.  For example,
\small
\begin{verbatim}
  call relax(U[device])  [[device]]
\end{verbatim}
\normalsize
executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
whereby the double square bracket notation indicates \emph{where} the function will execute.
Funtions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the
\texttt{value} attribute).

@@

In addition, CAFe subimage tasks may also be defined by the body of a Fortran do
concurrent construct and executed on a subimage.  For example, the code segment,
\small
\begin{verbatim}
do concurrent(i=1:N-1)  [[device]]
 begin block
   real :: w = 0.67
   S(i) = (1-w)*T(i) + w*(T(i-1)+T(i+1)/2)
 end block
end do
\end{verbatim}
\normalsize
will execute all of the code within the \texttt{do concurrent} construct on \texttt{device}.
Execution of a \texttt{do concurrent} loop on a subimage acts \emph{as if} the loop were
extracted as a outlined function and the subsequent task executed on a subimage.  Thus
there is an implicit barrier at the end of the loop that blocks the hosting image
from continuing execution until \emph{all} of the iterates of the loop have finished.

@@

Thus we introduce another CAFe function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
variable representing a fully-formed (established) team consisting of only one team
member, the current subimage.  If this function is called outside the context of a task
executing on a subimage it will return only a team consisting of the current image,
\texttt{this\_image()}.  Synchronization across possible subimage threads is then possible
using the statement
\small
\begin{verbatim}
  sync team (GET_SUBIMAGE_TEAM())
\end{verbatim}
\normalsize
This allows implementation of the technique of double buffering within a subimage task
so that all of the \emph{potential} subimage threads will have completed operating on the
temporary buffer before it is copied back to the primary array variable.

Fortran 2015 also introduces the concept of events.  An event is a Fortran type that
can be posted to by one image and waited for by another image.  CAFe relaxes the
constraint that events must be coarrays so that events can be used locally within
a single image to control concurrent execution of multiple subimage tasks.  For example,
the following code segment shows how two tasks can be run concurrently on two separate
subimages:
\small
\begin{verbatim}
 type(event_type) :: evt
 call update_ocn [[dev1, WITH_EVENT=evt]]
 call update_atm [[dev2, WITH_EVENT=evt]]
 event wait (evt, until_count=2)
\end{verbatim}
\normalsize
